NumPy---Array Programming in Python

### 2020-01-16 body.tex
Minor edits

Other utility functions include creating, reshaping, concatenating, and padding
arrays; searching, sorting and counting data; and reading and writing files.
NumPy provides extensive support for generating pseudorandom numbers and
includes an assortment of probability distributions.
It also performs accelerated linear algebra, utilizing one of several backends
such as OpenBLAS and Intel MKL optimized for the CPUs at hand.

@@

Various libraries build on and extend this functionality to further enrich the ecosystem of available tools.  For example, SciPy adds various mathematical, scientific, and engineering routines; Pandas provides indexed arrays, known as dataframes; xarray implements arrays with labeled axes; and Matplotlib generates publication-ready figures and visualizations.  Employing these low-level utilities are several widely used technique-specific libraries, including scikit-learn (machine learning), scikit-image (image processing), statmodels (statistics),  networkx (graphs and complex networks), and Napari (volumetric image visualization) \fixme{citations for all these}.

Finally, domain-specific libraries and applications are built on this diverse foundation of tooling; these include astropy (astronomy) \fixme{citation}, NiPy (neuroimaging) \fixme{citation}, and pangeo (earth science) \fixme{citation}.

% ecosystem
% core: scipy, pandas, matplolib ;
% scikits: scikit-learn, scikit-image, statmodels, networkx
% domain specific: astropy (2 citations?), neuroscience (3 citations?), pangeo, napari, etc.
% Ipython / Jupyter

@@

NumPy was initially developed by students, faculty, and researchers in their spare time to provide a modern array programming library for Python.  They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.  They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceeded Python like as Fortran \fixme{Cite https://www.netlib.org/} and C.  It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.

% Fortran / C
% IDL / FITS / ROOT / Parquet / Arrow ?

@@

This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.  For example, in a recent award given by the Royal Astronomical Society to AstroPy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists with experience in professional-standard software development practices including use of version control, unit testing, code review and issue tracking procedures. This is a vital skillset for modern researchers that is often missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education through formal courses and workshops\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}\fixme{cite https://scipy-school.org/}.

>>>>

Other utility functions include creating, reshaping, concatenating, and padding
arrays; searching, sorting and counting data; and reading and writing files.
NumPy provides extensive support for generating pseudorandom numbers and
includes an assortment of probability distributions.
It also performs accelerated linear algebra, utilizing one of several backends
such as OpenBLAS \cite{wang2013augem,xianyi2012model} and Intel MKL optimized for the CPUs at hand.

@@

Various libraries build on and extend this functionality to further enrich the
ecosystem of available tools.  For example, SciPy adds various mathematical,
scientific, and engineering routines; Pandas provides indexed arrays, known as
dataframes; xarray implements arrays with labeled axes; and Matplotlib
generates publication-ready figures and visualizations.  Employing these
low-level utilities are several widely used technique-specific libraries,
including scikit-learn (machine learning), scikit-image (image processing),
statmodels (statistics),  networkx (graphs and complex networks) \cite{SciPyProceedings_11}, and Napari
(volumetric image visualization).

Finally, domain-specific libraries and applications are built on this diverse
foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.

@@

NumPy was initially developed by students, faculty, and researchers in their spare time to provide a modern array programming library for Python.  They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.  They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceded Python like as Fortran \cite{dongarra2008netlib} and C.  It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.

% Fortran / C

@@

This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.  For example, in a recent award given by the Royal Astronomical Society to AstroPy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists with experience in professional-standard software development practices including use of version control, unit testing, code review and issue tracking procedures. This is a vital skillset for modern researchers that is often missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
%\fixme{cite https://scipy-school.org/}.
